Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-05-28 ThesisJP/Background/background.tex
Started integrated comments from sverre.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation were also growing until the required cooling finally became too much in 2004.  From that point, CPU manufactures had to look at different strategies than just increasing the clock frequency. One strategy was to embed several cores in one CPU. This ensured continuation of Moore's law, since the number of transistors per chip (CPU) could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs were typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation were also growing until the required cooling finally became too much in 2004.  From that point, CPU manufactures had to look at different strategies than just increasing the clock frequency. One strategy was to embed several cores in one CPU. This ensured continuation of Moore's law, since the number of transistors per chip (CPU) could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

### 2014-05-28 ThesisJP/Thesis.tex
Started integrated comments from sverre.

% ENGLISH ABSTRACT / SUMMARY
\CustChapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (GPU). This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed. When researchers are exploring new algorithms for ultrasound imaging, it is therefore important to keep the architecture of parallel accelerators like the GPU in mind. If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

In the first part of the thesis, the Capon adaptive beamformer is investigated and implemented on a GPU for the application of real-time sonar (\textbf{Paper\,I}) and medical ultrasound imaging (\textbf{Paper\,II}). Real-time frame rates are achieved for both modalities. \textbf{Paper\,II} also present, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images. In \textbf{Paper\,III}, we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting. A method is then proposed that improves the shift-invariant property. Shift-invariant imaging is essential if the method is ever to be used in practice.

In paper \textbf{Paper\,IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images. The method is capable of removing noise that by conventional methods would have occlude cardiac tissue. This work also shows that with modern GPUs it is possible to add advanced visualization methods to an ultrasound imaging system and still have real-time performance.

@@

% PREFACE AND ACKNOWLEDGEMENTS
\CustChapter*{Preface}%
\addcontentsline{toc}{chapter}{Preface}
%\thispagestyle{empty}%
This thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU). The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging. The main supervisor has been Professor Sverre Holm in the Department of Informatics, University of Oslo (UiO), and Professor II at MI-Lab, NTNU. Co-supervisors have been Ph.D. Stein Inge Rabben, Ph.D. Erik Steen, both at GE Vingmed Ultrasound AS, and Professor Hans Torp in the Department of Circulation and Medical Imaging, NTNU.

@@

During these years I have had the great pleasure of working with and getting to know several research communities and companies. I therefore would like to thank all colleagues at the Department of Circulation and Medical Imaging, NTNU, the Department of Informatics, UiO, GE Vingmed Ultrasound AS, and Squarehead Technology AS for providing great working environments. A special thanks goes to all my fellow Ph.D. students at MI-Lab. I will never forget our conference trips. A special thanks also goes to Jo Inge Buskenes at "office" 4470 for making every working hour so much fun.

@@

Finally I would like to thank my family and all my friends for their support, and for providing a world away from work. But most of all, I want to thank Beate for her love and patient. Without your support I would never have managed to complete this thesis.
\null
\vfill
\hfill Dedicated to Sofie Mathilde \AA{}sen
\CustChapter*{List of publications}
\addcontentsline{toc}{chapter}{List of publications}

>>>>

% ENGLISH ABSTRACT / SUMMARY
\CustChapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics processing unit (GPU). This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed. When researchers are exploring new algorithms for ultrasound imaging, it is therefore important to keep the architecture of parallel accelerators like the GPU in mind. If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

In the first part of the thesis, the Capon adaptive beamformer is investigated and implemented on a GPU for the application of real-time sonar (\textbf{Paper\,I}) and medical ultrasound imaging (\textbf{Paper\,II}). Real-time frame rates are achieved for both modalities. \textbf{Paper\,II} also presents, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images. In \textbf{Paper\,III}, we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting. A method is then proposed that improves the shift-invariant property. Shift-invariant imaging is essential if the method is ever to be used in practice.

In paper \textbf{Paper\,IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images. The method is capable of removing noise that by conventional methods would have occluded cardiac tissue. This work also shows that with modern GPUs it is possible to add advanced visualization methods to an ultrasound imaging system and still have real-time performance.

@@

% PREFACE AND ACKNOWLEDGEMENTS
\CustChapter*{Preface}%
\addcontentsline{toc}{chapter}{Preface}
%\thispagestyle{empty}%
This thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU). The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging. The main supervisor has been Professor Sverre Holm in the Department of Informatics (IFI), University of Oslo (UiO), and Professor II at MI-Lab, NTNU. Co-supervisors have been Ph.D. Stein Inge Rabben, Ph.D. Erik Steen, both at GE Vingmed Ultrasound AS, and Professor Hans Torp in the Department of Circulation and Medical Imaging, NTNU.

@@

During these years I have had the great pleasure of working with and getting to know several research communities and companies. I therefore would like to thank all colleagues at the Department of Circulation and Medical Imaging, NTNU, the Department of Informatics, UiO, GE Vingmed Ultrasound AS, and Squarehead Technology AS for providing great working environments. A special thanks goes to all my fellow Ph.D. students at MI-Lab. I will never forget our conference trips. A special thanks also goes to Jo Inge Buskenes in "office" 4470 at IFI, UiO for making every working hour so much fun.

@@

Finally I would like to thank my family and all my friends for being there, and for providing a world away from work. But most of all, I want to thank Beate for her love and patience. Without your support I would never have managed to complete this thesis.
\null
\vfill
\hfill Dedicated to Sofie Mathilde \AA{}sen
\CustChapter*{List of publications}
\addcontentsline{toc}{chapter}{List of publications}
